Lower Bounds on the Minimax Risk for the \\ Source Localization Problem

### 2017-05-08 paper.tex
WIP: Make Pulkit's edits; remove hyperlinks.

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{cite}
\usepackage[pdftex]{graphicx}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{mathtools}
\usepackage{nicefrac}
\usepackage{array}
\usepackage{enumitem}
\usepackage{color}
\usepackage{url}
\usepackage[colorlinks=true,allcolors=black,urlcolor=blue]{hyperref}

@@

\author{
	\IEEEauthorblockN{
		Praveen Venkatesh (\href{mailto:vpraveen@cmu.edu}{\texttt{vpraveen@cmu.edu}}) %\IEEEauthorrefmark{1}
		and Pulkit Grover (\href{mailto:pulkit@cmu.edu}{\texttt{pulkit@cmu.edu}}) %\IEEEauthorrefmark{2}
	}
	\IEEEauthorblockA{
		Electrical \& Computer Engineering,
		and the Center for the Neural Basis of Cognition,
		Carnegie Mellon University
		%\\
		%\IEEEauthorrefmark{1}\href{mailto:vpraveen@cmu.edu}{\texttt{vpraveen@cmu.edu}}
		%\IEEEauthorrefmark{2}\href{mailto:pulkit@cmu.edu}{\texttt{pulkit@cmu.edu}}
	}
	\thanks{This work was supported in part by Systems on Nanoscale
	Information fabriCs (SONIC), one of the six SRC STARnet Centers, sponsored
	by MARCO and DARPA.  Praveen Venkatesh was supported (in part) by the Dowd
	Fellowship from the College of Engineering at Carnegie Mellon University.
	The authors would like to thank Philip and Marsha Dowd for their financial
	support and encouragement. The authors thank Prof.\ Aarti Singh for her
	useful comments on a preliminary draft of this paper.
	}
	\vspace{-0.25in}
}

@@

The ``source localization'' problem is one in which we estimate the location of
a point source observed through a diffusive medium using an array of sensors.

We give lower bounds on the minimax risk (mean squared-error in location) in
estimating the location of the source, which apply to all estimators, for
certain classes of diffusive media, when using a uniformly distributed sensor
array. We show that for sensors of a fixed size, the lower bound decays with
increasing numbers of sensors. We also analyze a more physical sensor model to
understand the effect of shrinking the size of sensors as their number
increases to infinity, wherein the bound saturates for large sensor numbers.
In this scenario, it is seen that there is greater benefit to increasing the
number of sensors as the signal-to-noise ratio increases.  Our bounds are the
first to give a scaling for the minimax risk in terms of the number of sensors
used.

@@

The source localization problem arises in many fields in different forms. Our
principal motivation comes from Electroencephalography (EEG), a non-invasive
brain measurement modality which uses electrodes placed on the scalp to sense
electric potentials produced by neuronal activity within the
brain~\cite{Nunez2006Electric}. The neurons (or groups of neurons) which
produce this activity are usually modeled as current dipoles, and it is often
of clinical or neuroscientific interest to estimate their
locations~\cite{Baillet2001Electromagnetic}. For a single activated dipole
within the brain (which is modeled as a point source), the electrodes sample a
diffuse (blurred) representation of the dipole, as a result of the spatial
low-pass-filtering effect of the different layers of the skull, between the
brain and scalp.

Recent developments~\cite{Grover2016Information} have suggested that source
localization can benefit greatly from an increase in sensor density. This
conclusion is based primarily on a Nyquist rate analysis, and on an
information-theoretic bound~\cite{Grover2016Fundamental} on the accuracy of
source localization algorithms. This information-theoretic bound, however, was
derived for the unsampled, continuous-space potential on the scalp, or
equivalently, in the limit of the number of sensors going to infinity. The
claim that source localization can benefit from increased sensor density is,
therefore, not fully substantiated by the information-theoretic bound, since
the latter does not provide a lower bound showing improvement with increasing
sensor density.

Earlier work by Mosher et al.~\cite{Mosher1993Error} also gave numerical lower
bounds on source localization error. However, the Cramer-Rao lower bounds
derived by them do not apply to biased estimators, including several commonly
used source localization
algorithms~\cite{Hamalainen1994Interpreting,Lin2006Assessing}.

This paper seeks to address the shortcomings of earlier methods by deriving
lower bounds on the minimax risk for the source localization problem, which
apply to \emph{all} estimators (biased \emph{and} unbiased), and which capture
how the bound scales with increasing numbers of sensors. Previous lower bounds
are based on ``spherical head
models''~\cite{Nunez2006Electric,Grover2016Information}, which admit analytical
solutions for the potential of a single dipole. But spherical models are not
easily amenable to analytical lower bounds which seek to capture scaling in the
number of sensors, because the spherical surface does not permit uniform
sampling~\cite{Heath1956Thirteen}. We therefore restrict our analysis to a
source localization problem on a one-dimensional ``circular'' domain (to be
described in detail in Section~\ref{sec:source-localization}).  While this
simplified model makes crucial assumptions (such as shift-invariance) that are
invalid in more realistic brain models, we hope that the techniques presented
here will pave the way for deriving numerical bounds in more complex settings.

@@

Hence, we believe that this paper is the first to give lower bounds on the
minimax error in estimating the location of a point source observed by sensors
through a diffusive medium, and corrupted by additive white Gaussian noise. We
are also the first to analyze the implications of physical sensor models on
such lower bounds. A simplistic theoretical model might assume that the number
of sensors can increase unbounded, while maintaining a fixed SNR at each
sensor. We show that this is not physically possible in our setting, and derive
numerical bounds which have very different asymptotic behaviour.

@@

Plotting this numerically-computed bound for different numbers of trials
reveals some interesting trends (see Fig.~\ref{fig:numerical}). First, the
bound saturates as $m$ grows large, matching the behaviour of the
information-theoretic bound. Second, the lower bound on error is larger for
smaller numbers of sensors, as we expect. Third and most importantly, as the
number of trials increases, the bound begins to saturate only at a larger
number of sensors. In other words, there is \emph{increased} benefit in using
more sensors, when we \emph{also} have more trials. This substantiates the
intuition given in~\cite{Grover2016Information}, which makes this point using a
Nyquist-rate argument: a larger number of sensors will sample a larger number
of high-frequency components of the source signal (and hence allow us to locate
it better), but only if the noise floor is sufficiently low. Noise variance, in
turn, can be reduced by averaging over multiple trials.

>>>>

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{cite}
\usepackage[pdftex]{graphicx}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{mathtools}
\usepackage{nicefrac}
\usepackage{array}
\usepackage{enumitem}
\usepackage{color}
%\usepackage{url}
%\usepackage[colorlinks=true,allcolors=black,urlcolor=blue]{hyperref}

@@

\author{
	\IEEEauthorblockN{
		Praveen Venkatesh (\texttt{vpraveen@cmu.edu}) %(\href{mailto:vpraveen@cmu.edu}{\texttt{vpraveen@cmu.edu}}) %\IEEEauthorrefmark{1}
		and Pulkit Grover (\texttt{pulkit@cmu.edu}) %(\href{mailto:pulkit@cmu.edu}{\texttt{pulkit@cmu.edu}}) %\IEEEauthorrefmark{2}
	}
	\IEEEauthorblockA{
		Electrical \& Computer Engineering,
		and the Center for the Neural Basis of Cognition,
		Carnegie Mellon University
		%\\
		%\IEEEauthorrefmark{1}\href{mailto:vpraveen@cmu.edu}{\texttt{vpraveen@cmu.edu}}
		%\IEEEauthorrefmark{2}\href{mailto:pulkit@cmu.edu}{\texttt{pulkit@cmu.edu}}
	}
	\thanks{
	\textcolor{red}{The full version of this paper is available online~\cite{FullVersion}.}

	This work was supported in part by NSF CNS-1702694 and by Systems on
	Nanoscale Information fabriCs (SONIC), one of the six SRC STARnet Centers,
	sponsored by MARCO and DARPA.  Praveen Venkatesh was supported (in part) by
	the Dowd Fellowship from the College of Engineering at Carnegie Mellon
	University.  The authors would like to thank Philip and Marsha Dowd for
	their financial support and encouragement. The authors thank Prof.\ Anand
	Sarwate and Prof.\ Aarti Singh for their useful comments.
	}
	\vspace{-0.25in}
}

@@

The ``source localization'' problem is one in which we estimate the location of
a point source observed through a diffusive medium using an array of sensors.
We obtain lower bounds on the minimax risk (mean squared-error in location) in
estimating the location of the source, which apply to all estimators, for
certain classes of diffusive media, when using a uniformly distributed sensor
array. We show that for sensors of a fixed size, the lower bound decays to zero
with increasing numbers of sensors. We also analyze a more physical sensor
model to understand the effect of shrinking the size of sensors as their number
increases to infinity, wherein the bound saturates for large sensor numbers.
In this scenario, it is seen that there is greater benefit to increasing the
number of sensors as the signal-to-noise ratio increases.  Our bounds are the
first to give a scaling for the minimax risk in terms of the number of sensors
used.

@@

The source localization problem arises in many fields in different forms. Our
principal motivation comes from Electroencephalography (EEG), a non-invasive
brain measurement modality which uses electrodes placed on the scalp to sense
electric potentials produced by neuronal activity within the
brain~\cite{Nunez2006Electric}. The neurons (or groups of neurons) which
produce this activity are usually modeled as current dipoles, and it is often
of clinical or neuroscientific interest to estimate their
locations~\cite{Baillet2001Electromagnetic}. For a single activated dipole
within the brain (which is modeled as a point source), the electrodes sample a
diffuse (blurred) representation of the dipole, as a result of the spatial
low-pass-filtering effect of the different layers of the head, between the
brain and scalp.

\textcolor{red}{
\sout{Recent developments} Our recent work~\cite{Grover2016Information}
\sout{have suggested} suggests that source localization can benefit greatly
from an increase in sensor density. \sout{This conclusion is based primarily on
	a Nyquist rate analysis} Revisiting the Nyquist rate arguments presented
in~\cite{Nunez2006Electric} revealed that the question of how source imaging
resolution is affected by sensor density has not been addressed. We argued
intuitively that under appropriate conditions the right algorithms might be
able to recover the brain signal with much higher fidelity if more sensors were
used. To follow up on these intuitive arguments, Grover provides an
information-theoretic bound~\cite{Grover2016Fundamental} on the accuracy of
source localization algorithms.} This information-theoretic bound, however, was
derived for the unsampled, continuous-space potential on the scalp, or
equivalently, in the limit of the number of sensors going to infinity. \textcolor{red}{\sout{The
claim that source localization can benefit from increased sensor density is,
therefore, not fully substantiated by the information-theoretic bound, since it
does not provide a lower bound that shows improvement with increasing sensor
density.} Hence, this bound does not substantiate the claim that source
localization can benefit from increased sensor density.}

%\textcolor{blue}{[NEW VER.1]
%Our recent work~\cite{Grover2016Information} has suggested that EEG source
%localization can benefit greatly from an increase in sensor density. Prior
%work~\cite{Nunez2006Electric} relied on spatial Nyquist rate analyses to
%understand what sensor density is required to reconstruct the EEG signal on the
%\emph{surface of the scalp} to within a given mean squared error.  However,
%that line of research neglected to describe what sensor density is needed to
%recover the signal on the \emph{surface of the brain}; this second question
%also requires an understanding of the noise power spectral density of scalp
%EEG, which has been little-studied. Hence, it is still unknown what the maximum
%imaging resolution attainable with a given number of sensors is, and lower
%bounds providing this understanding do not yet exist.
%}

%\textcolor{green}{[NEW VER.2]
%Three questions naturally arise in such a setting: (1) What algorithms for
%source localization are good? (2) What is the best achievable imaging
%resolution? (3) How does sensor density affect resolution? The first question
%has been addressed by a vast literature on source localization
%algorithms~\cite{Baillet2001Electromagnetic}. The second question calls for
%lower bounds, of which there is as yet very little
%understanding~\cite{Grover2016Fundamental,Mosher1993Error}. The question of
%sensor density has received a lot of attention, but according to us, it still
%remains an open problem. \cite{Nunez2006Electric} uses Nyquist rate arguments
%to understand the number of sensors required to reconstruct the EEG signal on
%the \emph{surface of the scalp} to within a certain MSE. However, they fail to
%address the more relevant question of the number of sensors needed for
%\emph{source localization}, i.e., the number of sensors needed to recover the
%signal on the \emph{surface of the brain}. We show
%in~\cite{Grover2016Information} that this second question begs an understanding
%of the spatial noise power spectral density of scalp EEG, and that under
%appropriate conditions the right algorithms might be able to recover the brain
%signal with much higher fidelity if more sensors were used.
%\cite{Grover2016Information} does not, however, provide rigorous lower bounds
%to substantiate these observations.
%}

%\textcolor{red}{Recent work by Grover~\cite{Grover2016Fundamental} has sought to provide
%    analytical lower bounds on the accuracy of source localization algorithms using an information-theoretic method.
%    but did not provide a lower bound showing if and how this accuracy improves
%    with increasing sensor density.}
Earlier work by Mosher et al.~\cite{Mosher1993Error} also gave numerical lower
bounds on source localization error, with \emph{some} discussion on sensor
density. However, the Cramer-Rao lower bounds derived by them do not apply to
biased estimators, including several commonly used source localization
algorithms~\cite{Hamalainen1994Interpreting,Lin2006Assessing}.

This paper seeks to address the shortcomings of earlier methods by deriving
lower bounds on the minimax risk for the source localization problem, which
apply to \emph{all} estimators (biased \emph{and} unbiased), and which capture
how the bound scales with increasing numbers of sensors. Previous lower bounds
are based on ``spherical head
models''~\cite{Nunez2006Electric,Grover2016Information}, which admit analytical
solutions for the potential of a single dipole. But spherical models are not
easily amenable to analytical lower bounds which seek to capture scaling in the
number of sensors, because the spherical surface does not permit uniform
sampling~\cite{Heath1956Thirteen}. We therefore restrict our analysis to a
source localization problem on a one-dimensional ``circular'' domain (to be
described in detail in Section~\ref{sec:source-localization}). \textcolor{red}{While this
simplified model makes crucial assumptions (such as shift-invariance) that do
not hold in realistic brain models, the techniques presented here extend
naturally for deriving numerical bounds in more complex settings.}

@@

Hence, we believe that this paper is the first to give lower bounds on the
minimax error in estimating the location of a point source observed by sensors
through a diffusive medium, and corrupted by additive white Gaussian noise. We
are also the first to analyze the implications of physical sensor models on
such lower bounds. A simplistic theoretical model might assume that the number
of sensors can increase without bound, while still maintaining a fixed SNR at
each sensor. We show that this is non-physical in our setting, and derive
numerical bounds which have radically different asymptotic behaviour.

@@

Plotting this numerically-computed bound for different numbers of trials
reveals some interesting trends (see Fig.~\ref{fig:numerical}). First, the
bound saturates as $m$ grows large, matching the behaviour of the
information-theoretic bound. Second, the lower bound on error is larger for
smaller numbers of sensors, as we expect. Third and most importantly, as the
number of trials increases, the bound begins to saturate only at a larger
number of sensors. In other words, there is \emph{increased} benefit in using
more sensors, when we \emph{also} have more trials. This substantiates the
intuition given in~\cite{Grover2016Information}: a larger number of sensors
will sample a larger number of high-frequency components of the source signal
(and hence allow us to locate it better), but only if the noise floor is
sufficiently low. Noise variance, in turn, can be reduced by averaging over
multiple trials.
